---
title: "Wine Quality Analysis with Advanced ML Techniques & Visualizations"
author: "Surya"
date: "06-Sep-2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
start_time <- Sys.time()
```

# Wine Quality Analysis with Advanced ML Techniques & Visualizations
Welcome to this comprehensive analysis of wine quality using advanced machine learning techniques and visualizations in R! 🍷

---

## 1. Load Required Libraries

Before we can analyze wine, we need to load the necessary R packages. These packages provide tools for data manipulation (`tidyverse`), machine learning (`caret`, `randomForest`, `e1071`, `nnet`), visualization (`ggplot2`, `corrplot`), and data wrangling (`reshape2`). If any package is missing, it will be installed automatically.

```{r libraries}
required_libs <- c("tidyverse", "caret", "randomForest", "e1071", "nnet", "corrplot", "RCurl", "reshape2")
for (lib in required_libs) {
  if (!require(lib, character.only = TRUE)) {
    install.packages(lib, repos = "http://cran.us.r-project.org")
    library(lib, character.only = TRUE)
  }
}
```

---

## 2. Download and Prepare Wine Quality Data

We use two datasets from the UCI Machine Learning Repository: one for red wine and one for white wine. These datasets contain physicochemical properties and quality ratings for thousands of wines. We combine them into a single dataset and add a `wine_type` column to distinguish between red and white.

```{r data-download}
cat("Downloading wine quality datasets...\n")
options(timeout = 300)

red_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv"
red_wine_raw <- read.csv(red_url, sep = ";", header = TRUE)
red_wine_raw$wine_type <- "red"

white_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-white.csv"
white_wine_raw <- read.csv(white_url, sep = ";", header = TRUE)
white_wine_raw$wine_type <- "white"

wine_data <- rbind(red_wine_raw, white_wine_raw)
cat("Combined dataset size:", nrow(wine_data), "wines\n")
```

---

## 3. Data Preprocessing and Feature Engineering

To improve model performance and interpretability, we create new features based on domain knowledge and interactions between variables. We also categorize wine quality into "Low", "Medium", and "High" for classification, and remove any incomplete data.

```{r feature-engineering}
wine_data <- wine_data %>%
  mutate(
    quality_category = case_when(
      quality <= 5 ~ "Low",
      quality <= 7 ~ "Medium",
      TRUE ~ "High"
    ),
    quality_category = factor(quality_category, levels = c("Low", "Medium", "High")),
    wine_type = factor(wine_type),
    acid_ratio = fixed.acidity / volatile.acidity,
    sugar_alcohol_ratio = residual.sugar / alcohol,
    sulfur_ratio = free.sulfur.dioxide / total.sulfur.dioxide,
    total_acidity = fixed.acidity + volatile.acidity + citric.acid,
    alcohol_sugar_interaction = alcohol * residual.sugar,
    density_alcohol_ratio = density / alcohol,
    ph_acidity_balance = pH * total_acidity,
    preservation_index = total.sulfur.dioxide / density,
    quality_compounds = sulphates * citric.acid,
    alcohol_squared = alcohol^2,
    volatile_acidity_squared = volatile.acidity^2,
    wine_type_numeric = ifelse(wine_type == "red", 1, 0),
    red_wine_alcohol = wine_type_numeric * alcohol,
    white_wine_sugar = (1 - wine_type_numeric) * residual.sugar
  ) %>%
  na.omit()
```

---

## 4. Exploratory Analysis

Let's get a sense of our data: how many wines fall into each quality category, and how many are red vs. white? This helps us understand class balance and potential biases.

```{r exploratory}
cat("\nQuality distribution:\n")
print(table(wine_data$quality_category))
cat("\nWine type distribution:\n") 
print(table(wine_data$wine_type))
```

---

## 5. Visualizations

Visualizations help us see patterns and relationships in the data.  
- **Quality Distribution:** Shows class imbalance.
- **Correlation Heatmap:** Reveals relationships between chemical properties.
- **Alcohol Content by Quality:** Explores how alcohol relates to perceived quality.

```{r visualizations, fig.width=7, fig.height=5}
# Plot 1: Quality Distribution
p1 <- wine_data %>%
  ggplot(aes(x = quality_category, fill = quality_category)) +
  geom_bar(alpha = 0.8) +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Distribution of Wine Quality Categories",
       subtitle = "Most wines are mediocre - just like my cooking",
       x = "Quality Category", 
       y = "Number of Wines") +
  scale_fill_manual(values = c("Low" = "#FF6B6B", "Medium" = "#4ECDC4", "High" = "#45B7D1")) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12))
print(p1)

# Plot 2: Correlation Heatmap
cor_data <- wine_data %>%
  select(fixed.acidity, volatile.acidity, citric.acid, residual.sugar, 
         chlorides, free.sulfur.dioxide, total.sulfur.dioxide, 
         density, pH, sulphates, alcohol, quality) %>%
  cor()
cor_melted <- reshape2::melt(cor_data)
names(cor_melted) <- c("Var1", "Var2", "value")
p2 <- cor_melted %>%
  ggplot(aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab",
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        plot.title = element_text(size = 14, face = "bold")) +
  labs(title = "Wine Chemistry Relationships",
       subtitle = "More complex than a sommelier's tasting notes",
       x = "", y = "") +
  coord_fixed()
print(p2)

# Plot 3: Alcohol Content by Quality
p3 <- wine_data %>%
  ggplot(aes(x = quality_category, y = alcohol, fill = quality_category)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.5) +
  geom_jitter(alpha = 0.3, width = 0.2, size = 0.5) +
  labs(title = "Alcohol Content vs Wine Quality",
       subtitle = "Higher alcohol = Better wine. I don't make the rules.",
       x = "Quality Category",
       y = "Alcohol Content (%)") +
  scale_fill_manual(values = c("Low" = "#FF6B6B", "Medium" = "#4ECDC4", "High" = "#45B7D1")) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12))
print(p3)
```

---

## 6. Train/Test/Validation Split

To evaluate our models fairly, we split the data into training, validation, and test sets.  
- **Training:** Used to fit models.
- **Validation:** Used for model selection and tuning.
- **Test:** Used for final unbiased evaluation.

```{r split}
set.seed(42)
test_index <- createDataPartition(y = wine_data$quality_category, times = 1, p = 0.2, list = FALSE)
train_set <- wine_data[-test_index,]
test_set <- wine_data[test_index,]

set.seed(42)
val_index <- createDataPartition(y = train_set$quality_category, times = 1, p = 0.2, list = FALSE)
validation_set <- train_set[val_index,]
train_set_final <- train_set[-val_index,]

cat("Training set size:", nrow(train_set_final), "\n")
cat("Validation set size:", nrow(validation_set), "\n") 
cat("Test set size:", nrow(test_set), "\n")
```

---

## 7. Model Training & Validation

We train several models to predict wine quality:
- **Baseline:** Always predicts the most common class.
- **Random Forest:** Ensemble of decision trees.
- **SVM:** Finds optimal boundaries between classes.
- **Neural Network:** Captures complex nonlinear relationships.
- **Enhanced Random Forest:** More trees and features for improved accuracy.

We compare their validation accuracy to select the best approach.

```{r model-training}
accuracy_func <- function(actual, predicted) {
  mean(actual == predicted)
}

feature_cols <- c("fixed.acidity", "volatile.acidity", "citric.acid", "residual.sugar",
                  "chlorides", "free.sulfur.dioxide", "total.sulfur.dioxide", "density", 
                  "pH", "sulphates", "alcohol", "wine_type", "acid_ratio", 
                  "sugar_alcohol_ratio", "sulfur_ratio", "total_acidity",
                  "alcohol_sugar_interaction", "density_alcohol_ratio", "ph_acidity_balance",
                  "preservation_index", "quality_compounds", "alcohol_squared",
                  "volatile_acidity_squared", "red_wine_alcohol", "white_wine_sugar")

results_df <- data.frame(Method = character(), Accuracy = numeric(), stringsAsFactors = FALSE)

# Baseline
baseline_pred <- rep(names(sort(table(train_set_final$quality_category), decreasing = TRUE))[1], 
                     nrow(validation_set))
baseline_accuracy <- accuracy_func(validation_set$quality_category, baseline_pred)
results_df <- rbind(results_df, data.frame(Method = "Baseline (Most Frequent)", Accuracy = baseline_accuracy))
cat("Method 1 completed: Baseline\n")

# Random Forest
set.seed(42)
rf_model <- randomForest(quality_category ~ ., 
                         data = train_set_final[, c(feature_cols, "quality_category")],
                         ntree = 500, mtry = 8, importance = TRUE)
rf_pred <- predict(rf_model, validation_set)
rf_accuracy <- accuracy_func(validation_set$quality_category, rf_pred)
results_df <- rbind(results_df, data.frame(Method = "Random Forest", Accuracy = rf_accuracy))
cat("Method 2 completed: Random Forest\n")

# SVM
set.seed(42)
svm_model <- svm(quality_category ~ ., 
                 data = train_set_final[, c(feature_cols, "quality_category")],
                 kernel = "radial", cost = 1, gamma = 0.1, probability = TRUE)
svm_pred <- predict(svm_model, validation_set)
svm_accuracy <- accuracy_func(validation_set$quality_category, svm_pred)
results_df <- rbind(results_df, data.frame(Method = "SVM (RBF Kernel)", Accuracy = svm_accuracy))
cat("Method 3 completed: SVM\n")

# Neural Network
set.seed(42)
numeric_cols <- c("fixed.acidity", "volatile.acidity", "citric.acid", "residual.sugar",
                  "chlorides", "free.sulfur.dioxide", "total.sulfur.dioxide", "density", 
                  "pH", "sulphates", "alcohol", "acid_ratio", "sugar_alcohol_ratio", 
                  "sulfur_ratio", "total_acidity", "alcohol_sugar_interaction",
                  "density_alcohol_ratio", "ph_acidity_balance", "preservation_index",
                  "quality_compounds", "alcohol_squared", "volatile_acidity_squared",
                  "red_wine_alcohol", "white_wine_sugar")
train_scaled <- train_set_final
validation_scaled <- validation_set
for(col in numeric_cols) {
  col_mean <- mean(train_set_final[[col]])
  col_sd <- sd(train_set_final[[col]])
  train_scaled[[col]] <- (train_set_final[[col]] - col_mean) / col_sd
  validation_scaled[[col]] <- (validation_set[[col]] - col_mean) / col_sd
}
nn_model <- nnet(quality_category ~ ., 
                 data = train_scaled[, c(feature_cols, "quality_category")],
                 size = 15, decay = 0.1, maxit = 300, trace = FALSE)
nn_pred <- predict(nn_model, validation_scaled, type = "class")
nn_accuracy <- accuracy_func(validation_set$quality_category, nn_pred)
results_df <- rbind(results_df, data.frame(Method = "Neural Network", Accuracy = nn_accuracy))
cat("Method 4 completed: Neural Network\n")

# Enhanced Random Forest
set.seed(42)
tuned_rf <- randomForest(quality_category ~ ., 
                         data = train_set_final[, c(feature_cols, "quality_category")],
                         ntree = 1000, mtry = 10, importance = TRUE, nodesize = 2)
tuned_rf_pred <- predict(tuned_rf, validation_set)
tuned_rf_accuracy <- accuracy_func(validation_set$quality_category, tuned_rf_pred)
results_df <- rbind(results_df, data.frame(Method = "Enhanced Random Forest", Accuracy = tuned_rf_accuracy))
cat("Method 5 completed: Enhanced Random Forest\n")
```

---

## 8. Weighted Ensemble Model

Ensembles combine predictions from multiple models to improve accuracy and robustness. Here, we weight each model's predictions by its validation accuracy, giving more influence to better-performing models.

```{r ensemble}
cat("Creating Weighted Ensemble - combining the wisdom of all our models...\n")
set.seed(42)
rf_probs <- predict(rf_model, validation_set, type = "prob")
svm_probs <- attr(predict(svm_model, validation_set, probability = TRUE), "probabilities")
nn_probs <- predict(nn_model, validation_scaled, type = "raw")
tuned_rf_probs <- predict(tuned_rf, validation_set, type = "prob")
model_accuracies <- c(rf_accuracy, svm_accuracy, nn_accuracy, tuned_rf_accuracy)
ensemble_weights <- model_accuracies / sum(model_accuracies)
cat("Ensemble weights based on performance:\n")
cat("Random Forest:", round(ensemble_weights[1], 3), "\n")
cat("SVM:", round(ensemble_weights[2], 3), "\n")
cat("Neural Network:", round(ensemble_weights[3], 3), "\n")
cat("Enhanced RF:", round(ensemble_weights[4], 3), "\n")
weighted_probs <- ensemble_weights[1] * rf_probs + 
  ensemble_weights[2] * svm_probs[,c("Low","Medium","High")] + 
  ensemble_weights[3] * nn_probs + 
  ensemble_weights[4] * tuned_rf_probs
ensemble_pred <- factor(colnames(weighted_probs)[apply(weighted_probs, 1, which.max)], 
                        levels = c("Low", "Medium", "High"))
ensemble_accuracy <- accuracy_func(validation_set$quality_category, ensemble_pred)
results_df <- rbind(results_df, data.frame(Method = "Weighted Ensemble", Accuracy = ensemble_accuracy))
cat("Method 6 completed: Weighted Ensemble (our potential 90%+ superstar!)\n")
```

---

## 9. Feature Importance Visualization

Which features matter most for predicting wine quality? We visualize the top contributors according to our best-performing model, helping us understand what drives wine ratings.

```{r feature-importance, fig.width=7, fig.height=5}
best_model_idx <- which.max(results_df$Accuracy[1:6])
if(results_df$Method[best_model_idx] == "Weighted Ensemble") {
  importance_df <- data.frame(
    Feature = names(importance(tuned_rf)[, "MeanDecreaseGini"]),
    Importance = importance(tuned_rf)[, "MeanDecreaseGini"]
  ) %>%
    arrange(desc(Importance)) %>%
    slice_head(n = 10) %>%
    mutate(Feature = reorder(Feature, Importance))
  subtitle_text <- "Ensemble champion reveals the wine secrets (our features are legends!)"
} else {
  importance_df <- data.frame(
    Feature = names(importance(tuned_rf)[, "MeanDecreaseGini"]),
    Importance = importance(tuned_rf)[, "MeanDecreaseGini"]
  ) %>%
    arrange(desc(Importance)) %>%
    slice_head(n = 10) %>%
    mutate(Feature = reorder(Feature, Importance))
  subtitle_text <- "Enhanced Random Forest picks its favorites (engineered features FTW!)"
}
p4 <- importance_df %>%
  ggplot(aes(x = Feature, y = Importance, fill = Importance)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  scale_fill_gradient(low = "#E8F4F8", high = "#2E86AB") +
  labs(title = "Feature Importance: What Actually Predicts Wine Quality",
       subtitle = subtitle_text,
       x = "Features (Original + Our Engineered Ones)",
       y = "Importance Score") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12))
print(p4)
```

---

## 10. Model Performance Comparison

How do our models stack up? This section compares validation accuracy across all approaches, highlighting the strengths of ensemble learning.

```{r performance-comparison, fig.width=7, fig.height=5}
performance_df <- results_df[1:6, ]
p5 <- performance_df %>%
  ggplot(aes(x = reorder(Method, Accuracy), y = Accuracy, fill = Accuracy)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = round(Accuracy, 3)), hjust = -0.1) +
  coord_flip() +
  scale_fill_gradient(low = "#FFE5E5", high = "#2E8B57") +
  labs(title = "Model Performance Championship",
       subtitle = "Can our Ensemble break the 90% barrier? 🏆",
       x = "Machine Learning Models",
       y = "Validation Accuracy") +
  ylim(0, 1) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12))
print(p5)
```

---

## 11. Final Test Evaluation

We apply our best model to the held-out test set for an unbiased estimate of real-world performance. This is the ultimate test of our wine-judging AI!

```{r final-test}
best_model_idx <- which.max(results_df$Accuracy)
best_method <- results_df$Method[best_model_idx]
cat("\nBest validation method:", best_method, "\n")
cat("Validation accuracy:", round(results_df$Accuracy[best_model_idx], 4), "\n")

if(best_method == "Weighted Ensemble") {
  test_scaled <- test_set
  for(col in numeric_cols) {
    col_mean <- mean(train_set_final[[col]])
    col_sd <- sd(train_set_final[[col]])
    test_scaled[[col]] <- (test_set[[col]] - col_mean) / col_sd
  }
  rf_test_probs <- predict(rf_model, test_set, type = "prob")
  svm_test_probs <- attr(predict(svm_model, test_set, probability = TRUE), "probabilities")
  nn_test_probs <- predict(nn_model, test_scaled, type = "raw")
  tuned_rf_test_probs <- predict(tuned_rf, test_set, type = "prob")
  weighted_test_probs <- ensemble_weights[1] * rf_test_probs + 
    ensemble_weights[2] * svm_test_probs[,c("Low","Medium","High")] + 
    ensemble_weights[3] * nn_test_probs + 
    ensemble_weights[4] * tuned_rf_test_probs
  final_pred <- factor(colnames(weighted_test_probs)[apply(weighted_test_probs, 1, which.max)], 
                       levels = c("Low", "Medium", "High"))
} else if(grepl("Enhanced Random Forest", best_method)) {
  final_pred <- predict(tuned_rf, test_set)
} else if(best_method == "Random Forest") {
  final_pred <- predict(rf_model, test_set)
} else if(best_method == "SVM (RBF Kernel)") {
  final_pred <- predict(svm_model, test_set)
} else {
  test_scaled <- test_set
  for(col in numeric_cols) {
    col_mean <- mean(train_set_final[[col]])
    col_sd <- sd(train_set_final[[col]])
    test_scaled[[col]] <- (test_set[[col]] - col_mean) / col_sd
  }
  final_pred <- predict(nn_model, test_scaled, type = "class")
}
final_accuracy <- accuracy_func(test_set$quality_category, final_pred)
results_df <- rbind(results_df, data.frame(Method = "Final Test Accuracy", Accuracy = final_accuracy))
cat("\n=== RESULTS SUMMARY ===\n")
cat("Drumroll please... 🥁\n")
print(results_df)
if(final_accuracy >= 0.90) {
  cat("\n🎉 MISSION ACCOMPLISHED! We broke the 90% barrier! 🎉\n")
  cat("Our ensemble wine-judging AI is now better than most humans! 🍷\n")
} else {
  cat(sprintf("\n📈 Final accuracy: %.1f%% - Getting close to sommelier level!\n", final_accuracy * 100))
  cat("Still better than my wine-picking skills at the grocery store! 🛒\n")
}
```

---

## 12. Feature Importance & Confusion Matrix

To further interpret our model, we show the top features and a confusion matrix, plus precision, recall, and F1-score for each class. This gives a detailed view of strengths and weaknesses.

```{r importance-confusion}
cat("\nTop 10 Most Important Features:\n")
cat("AKA: What makes wine worth the price tag\n")
importance_scores <- importance(tuned_rf)[, "MeanDecreaseGini"]
top_features <- sort(importance_scores, decreasing = TRUE)[1:10]
print(round(top_features, 2))

cat("\nFinal Test Set Confusion Matrix:\n")
cat("Where our ensemble gets confused (hopefully rarely!)\n")
conf_matrix <- table(Predicted = final_pred, Actual = test_set$quality_category)
print(conf_matrix)

precision <- diag(conf_matrix) / rowSums(conf_matrix)
recall <- diag(conf_matrix) / colSums(conf_matrix)
f1 <- 2 * (precision * recall) / (precision + recall)

cat("\nPer-class Performance:\n")
cat("More detailed than a sommelier's tasting notes (and hopefully more accurate!)\n")
performance_summary <- data.frame(
  Class = names(precision),
  Precision = round(precision, 3),
  Recall = round(recall, 3),
  F1_Score = round(f1, 3)
)
print(performance_summary)
```

---

## 13. Execution Time Tracking

Finally, we report how long the analysis took. Good wine takes time, but good code should be efficient!

```{r timing}
end_time <- Sys.time()
total_time <- as.numeric(difftime(end_time, start_time, units = "mins"))
cat("\nTotal execution time:", round(total_time, 2), "minutes\n")
cat("Still faster than waiting for your Uber after wine tasting! 🍷\n")
cat("\n🎉 Analysis complete! Our weighted ensemble can now judge wine better than that friend who only drinks box wine but claims to be an expert! 🎉\n")
cat("With our ensemble approach combining multiple models, we're targeting true sommelier-level